---
title: "DataExploration"
output: html_document
---



```{r}
library(tidyverse)
regime.table <- read_rds('Data.rds')
```

# Year correlation
```{r}
regime.numeric <- regime.table %>% mutate(Regime = as.numeric(Regime) - 11)

plot(log(GDPPerCapita10) ~ Year, data = regime.numeric)
abline(reg = lm(log(GDPPerCapita10) ~ Year, data = regime.numeric))

plot(jitter(GiniCoef10) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(GiniCoef10) ~ Year, data = regime.numeric))

plot(jitter(DecadePoverty) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(DecadePoverty) ~ Year, data = regime.numeric))

plot(jitter(InfantMortality10) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(InfantMortality10) ~ Year, data = regime.numeric))

plot(log(Population10) ~ Year, data = regime.numeric)
abline(reg = lm(log(Population10) ~ Year, data = regime.numeric))

plot(log(YearsSchooling6.10) ~ Year, data = regime.numeric)
abline(reg = lm(log(YearsSchooling6.10) ~ Year, data = regime.numeric))
```


# GDP Correlation
```{r}
regime.numeric <- regime.table %>% mutate(Regime = as.numeric(Regime) - 11)

plot(jitter(GiniCoef10) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(GiniCoef10) ~ log(GDPPerCapita10), data = regime.numeric))

plot(jitter(DecadePoverty) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(DecadePoverty) ~ log(GDPPerCapita10), data = regime.numeric))

plot(jitter(InfantMortality10) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(InfantMortality10) ~ log(GDPPerCapita10), data = regime.numeric))

plot(log(Population10) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(log(Population10) ~ log(GDPPerCapita10), data = regime.numeric))

plot(YearsSchooling6.10 ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(YearsSchooling6.10 ~ log(GDPPerCapita10), data = regime.numeric))
```

# Regime
```{r}
plot(jitter(Regime) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ Year, data = regime.numeric))

plot(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric))

plot(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric))

plot(jitter(Regime) ~ DecadePoverty, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ DecadePoverty, data = regime.numeric))

plot(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric))

plot(jitter(Regime) ~ log(Population10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(Population10), data = regime.numeric))

plot(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric))
```


# Regime > 0
```{r}
regime.numeric <- regime.table %>% mutate(Regime = as.numeric(Regime) - 11)
regime.numeric <- regime.numeric[regime.numeric$Regime > 0,]

plot(jitter(Regime) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ Year, data = regime.numeric))

plot(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric))

plot(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric))

plot(jitter(Regime) ~ DecadePoverty, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ DecadePoverty, data = regime.numeric))

plot(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric))

plot(jitter(Regime) ~ log(Population10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(Population10), data = regime.numeric))

plot(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric))
```


# Regime <= 0
```{r}
regime.numeric <- regime.table %>% mutate(Regime = as.numeric(Regime) - 11)
regime.numeric <- regime.numeric[regime.numeric$Regime <= 0,]

plot(jitter(Regime) ~ Year, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ Year, data = regime.numeric))

plot(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(GDPPerCapita10), data = regime.numeric))

plot(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(GiniCoef10), data = regime.numeric))

plot(jitter(Regime) ~ DecadePoverty, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ DecadePoverty, data = regime.numeric))

plot(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ jitter(InfantMortality10), data = regime.numeric))

plot(jitter(Regime) ~ log(Population10), data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ log(Population10), data = regime.numeric))

plot(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric)
abline(reg = lm(jitter(Regime) ~ YearsSchooling6.10, data = regime.numeric))
```













